Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-25 paper.tex
runtime figures

\newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
% @plr - purple doesn't exist in my Latex
\newcommand{\plrcomment}[1]{\textcolor{blue}{#1}}

@@

\begin{figure}
% \begin{center}
% \begin{tabular}{cc}
% % \includegraphics[width=5cm]{msprime-scaling.png} &
% % \includegraphics[width=5cm]{figures/recombination} \\
% \end{tabular}
% \end{center}
% \includegraphics[width=\textwidth]{figures/recombination}
\Large{Plot of msprime run time against $\rho$}
\caption{\label{fig-recombination-perf} Approximate run time for \msprime\
simulations as a function of scaled recombination rate.
}
\end{figure}

@@

% \jkcomment{We want to round out this discussion here by giving some
% sort of rough argument for why Hudson's algorithm is quadratic, and discussing
% to Fig.~\ref{fig-recombination-perf} to show roughly what we can expect
% to simulate in how long. See issue 28.}

Figure~\ref{fig-recombination-perf} gives a rough idea of the runtime of
a given simulation of a constant size population.
However, what about a simulation with \emph{changing} size?
To understand how runtime depends on demography
it helps to consider \emph{why} runtime is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancesty,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid yet inconsequential events that lead to the quadratic runtime.
\plrcomment{(this is discussed also in the ARG section; check for overlap)}
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of runtime we can estimate when this occurs.
Suppose that the maximal number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
It is known that the maximum number of lineages is reached very quickly \citep{nelson2020accounting},
and at this maximum, recombination and coalescence rates are equal,
so many (roughly half) of these lineages have at least two distant segments of ancestral material
separated by a long segment of genome, implying that $\ell \approx L / 4$.
This implies that $M$ is roughly $L N_e$.
Now, the number of linages decreases gradually, on the coalescent time scale -- so, over roughly $2 N_e$ generations.
Since the total rate of events when there are the maximum number of lineages
happens at rate roughly $L^2 N_e / 4$,
then the total number of events is proportional to $(L N_e)^2$ -- i.e., proportional to $\rho^2$,
as desired.
What does this tell us about time-varying population sizes?
The argument above implies that the work is spread out relatively evenly on the coalescent time scale:
in other words,
more events happen in the time period from 0 to $N_e/2$ generations
than from $N_e/2$ to $N_e$ generations,
but the runtime is not dominated but the more recent time period.

Putting this together:
suppose that population size today is $N_1$,
while longer ago than $T$ generations ago it was $N_2$:
does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
The answer depends on how $T$ commpares to $N_1$: if $T/N_1$ is large,
then runtime will be similar to a population of size $N_1$;
while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
and the runtime will depend critically on $T$ -- in other words,
simulation will be quick in a species with a strong domestication bottleneck,
and slow otherwise.

>>>>

\newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
\newcommand{\plrcomment}[1]{\textcolor{cyan}{#1}}

@@

\begin{figure}
\begin{center}
    \includegraphics{figures/ancestry-perf}
\end{center}
\caption{\label{fig-ancestry-perf}
    Approximate run time for \msprime\ simulations,
    for \textbf{(left)} small, and \textbf{(right)} larger simulations.
    Each point is the runtime of one simulation,
    for various values of effective population size ($N$),
    genome length ($L$), and number of samples.
    Runtime scales quadratically with the product of $N$ and $L$;
    since recombination rate in these simulations was $10^{-8}$,
    the scaled recombination rate here is $\rho = NL/10^8$,
    shown on the horizontal axis.
    The black line is a quadratic fit separately in each panel.
    \plrcomment{TODO: run out more simulations to get more triangles}
}
\end{figure}

@@

% \jkcomment{We want to round out this discussion here by giving some
% sort of rough argument for why Hudson's algorithm is quadratic, and discussing
% to Fig.~\ref{fig-ancestry-perf} to show roughly what we can expect
% to simulate in how long. See issue 28.}

Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
a given simulation of a constant size population,
and verifies that runtime scales quadratically with $\rho$,
and depends much less strongly on the number of samples.
Taking a typical chromosome to be 1 Morgan in length,
these plot show, roughly, that simulating
samples from a population of thousands of individuals takes seconds,
while samples from a population of tens of thousands takes minutes.

However, what about a simulation with \emph{changing} size?
To understand how runtime depends on demography
it helps to consider \emph{why} runtime is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancesty,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid yet inconsequential events that lead to the quadratic runtime.
\plrcomment{(this is discussed also in the ARG section; check for overlap)}
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of runtime we can estimate when this occurs.
Suppose that the maximal number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
It is known that the maximum number of lineages is reached very quickly \citep{nelson2020accounting},
and at this maximum, recombination and coalescence rates are equal,
so many (roughly half) of these lineages have at least two distant segments of ancestral material
separated by a long segment of genome, implying that $\ell \approx L / 4$.
This implies that $M$ is roughly $L N_e$.
Now, the number of linages decreases gradually, on the coalescent time scale -- so, over roughly $2 N_e$ generations.
Since the total rate of events when there are the maximum number of lineages
happens at rate roughly $L^2 N_e / 4$,
then the total number of events is proportional to $(L N_e)^2$ -- i.e., proportional to $\rho^2$,
as desired.
What does this tell us about time-varying population sizes?
The argument above implies that the work is spread out relatively evenly on the coalescent time scale:
in other words,
more events happen in the time period from 0 to $N_e/2$ generations
than from $N_e/2$ to $N_e$ generations,
but the runtime is not dominated but the more recent time period.

Putting this together:
suppose that population size today is $N_1$,
while longer ago than $T$ generations ago it was $N_2$:
does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
The answer depends on how $T$ commpares to $N_1$: if $T/N_1$ is large,
then runtime will be similar to a population of size $N_1$;
while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
and the runtime will depend critically on $T$ -- in other words,
simulation will be quick in a species with a strong domestication bottleneck,
and slow otherwise.
In practice, we recommend that users estimate runtime by
measuring runtime in a series of simulations with short genome lengths,
possibly varying other parameters to check for interactions,
and then predict runtime by fitting a quadratic curve to genome length.
